---
title: "CS_02_BELLABEAT"
author: "RJLD"
date: "2025-04-08"
output: html_document
---

<!-- my dataset covers 4.12.16 to 5.12.16, sleep data is 4.12.16 to 5.12.16. These in csv format-->

#INSTALL & LOAD PACKAGES
Install the following packages first before running <!--removing code chunk for installpackageds coz can't knit-->
install.packages("tidyverse")
install.packages("here")
install.packages("janitor")
install.packages("skimr")
install.packages("dplyr")
install.packages("lubridate")
install.packages("hms")



```{r}
library(tidyverse)
library(here)
library(janitor)
library(skimr)
library(dplyr)
library(lubridate)
library(hms)
```

#LOAD DATASET & DATA EXPLORE & VISUALIZE
##DAILY
###DAILY-LOAD
```{r}
daily_activity_A <- read.csv("D:/Users/rjlde/Documents/+STUDY FILES+/2025 CASE STUDIES/CASE STUDY 2/RPROJ/CSV/DAILY/dailyActivity_merged_3.12.16-4.11.16.csv", stringsAsFactors = FALSE)
daily_activity_B <- read.csv("D:/Users/rjlde/Documents/+STUDY FILES+/2025 CASE STUDIES/CASE STUDY 2/RPROJ/CSV/DAILY/dailyActivity_merged_4.12.16-5.12.16.csv", stringsAsFactors = FALSE)
sleep_day_B <- read.csv("D:/Users/rjlde/Documents/+STUDY FILES+/2025 CASE STUDIES/CASE STUDY 2/RPROJ/CSV/DAILY/sleepDay_merged_4.12.16-5.12.16.csv", stringsAsFactors = FALSE)
```
###DAILY-CLEAN
head()
```{r}
head(daily_activity_A)
head(daily_activity_B)
head(sleep_day_B)#replace daily_activity_A w/ daily_activity_B and w/ sleep_day_B
#Findings - ActivityDate is chr not datetime format
```
Number of Unique participants:
```{r}
n_distinct(daily_activity_A)#replace daily_activity_A w/ daily_activity_B and w/ sleep_day_B
nrow(daily_activity_A)
```
####Clean ->
remove duplicated rows:
```{r}
daily_activity_A <- daily_activity_A[!duplicated(daily_activity_A), ]
daily_activity_B <- daily_activity_B[!duplicated(daily_activity_B), ]
sleep_day_B <- sleep_day_B[!duplicated(sleep_day_B), ]
#no duplicated values found
```
Handle missing values:
```{r}
colSums(is.na(daily_activity_A))
colSums(is.na(daily_activity_B))
colSums(is.na(sleep_day_B))# replace a,b, sleep
#no missing values found
```
Convert data types:
```{r}
daily_activity_A$ActivityDate <- as.Date(daily_activity_A$ActivityDate, format = "%m/%d/%Y")
daily_activity_B$ActivityDate <- as.Date(daily_activity_B$ActivityDate, format = "%m/%d/%Y")
sleep_day_B$SleepDay <- as.Date(sleep_day_B$SleepDay, format = "%m/%d/%Y %H:%M:%S" )
#converted chr to date
```

Quick Summary Stats:
```{r}
summary_daily_activity_A <- daily_activity_A %>%
    summary()

summary_daily_activity_B <-daily_activity_B %>% 
    summary()

summary_sleep_day_B <-sleep_day_B %>%
  select(TotalSleepRecords,
  TotalMinutesAsleep,
  TotalTimeInBed) %>%
    summary()
```
Save summary stats
```{r}
write.csv(summary_daily_activity_A, "D:/Users/rjlde/Documents/+STUDY FILES+/2025 CASE STUDIES/CASE STUDY 2/RPROJ/CSV/CLEANED/summary_daily_activity_A.csv")
write.csv(summary_daily_activity_B, "D:/Users/rjlde/Documents/+STUDY FILES+/2025 CASE STUDIES/CASE STUDY 2/RPROJ/CSV/CLEANED/summary_daily_activity_B.csv")
write.csv(summary_sleep_day_B, "D:/Users/rjlde/Documents/+STUDY FILES+/2025 CASE STUDIES/CASE STUDY 2/RPROJ/CSV/CLEANED/summary_sleep_day_B.csv")
```
View summary stats
```{r}
print(summary_sleep_day_B) #replace with wahat u want to view
```


####SAVE CLEANED FILES
```{r}
write.csv(daily_activity_A, "D:/Users/rjlde/Documents/+STUDY FILES+/2025 CASE STUDIES/CASE STUDY 2/RPROJ/CSV/CLEANED/daily_activity_A.csv", row.names = FALSE)
write.csv(daily_activity_B, "D:/Users/rjlde/Documents/+STUDY FILES+/2025 CASE STUDIES/CASE STUDY 2/RPROJ/CSV/CLEANED/daily_activity_B.csv", row.names = FALSE)
write.csv(daily_activity_A, "D:/Users/rjlde/Documents/+STUDY FILES+/2025 CASE STUDIES/CASE STUDY 2/RPROJ/CSV/CLEANED/sleep_day_B.csv", row.names = FALSE)
```


###DAILY-EXPLORATION & VISUALIZATION
####Relationship bet steps taken in a day and sedentary minutes:
```{r}
ggplot(data = daily_activity_B, aes(x = TotalSteps, y = SedentaryMinutes, color = factor(Id))) + #replace with A and caption diff
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "black") +  # Adds a linear regression line
 # facet_wrap(~ Id, scales = "fixed") +  #add this line if gusto hiwahiwalay per unique id
  labs(
    title = "Steps Taken Daily vs. Sedentary Minutes",
    caption = "Period: 04/12/2016 to 05/12/2016 Daily" ) + #if B
    #caption = "Period: 03/11/2016 to 04/12/2016 Daily") + #if A
  theme_minimal()
```
####Relationship bet. minutesofsleep & time in bed:
```{r}
ggplot(data=sleep_day_B, aes(x=TotalMinutesAsleep, y=TotalTimeInBed, color=factor(Id))) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "black") +  # Adds a linear regression line
  #facet_wrap(~ Id, scales = "fixed") +  #add this line if gusto hiwahiwalay per unique id
  labs(
    title="Total Minutes Asleep vs Total Time(minutes) in Bed",
    caption="Period: 04/12/2016 to 05/12/2016 Daily"
  ) +
  theme_minimal()
```

###DAILY-MERGE
```{r}
daily_merged_B <- merge(sleep_day_B, 
                        daily_activity_B, 
                        by.x = c("Id", "SleepDay"),  # Columns in sleep_day_B
                        by.y = c("Id", "ActivityDate"),  # Columns in daily_activity_B
                        all = TRUE)  # Keep all rows from both data frames

daily_merged_B <- daily_merged_B %>%
  rename(Date = SleepDay)

write.csv(daily_merged_B, "D:/Users/rjlde/Documents/+STUDY FILES+/2025 CASE STUDIES/CASE STUDY 2/RPROJ/CSV/CLEANED/daily_merged_B.csv", row.names = FALSE)
```
####check new table
```{r}
head(daily_merged_B)
nrow(daily_merged_B)
n_distinct(daily_merged_B)
```

####Relationship betweeen total steps and total minutes asleep
```{r}
ggplot(data=daily_merged_B, aes(x=TotalMinutesAsleep, y=TotalSteps, color=factor(Id)))+
  geom_point()+
  geom_smooth(method = "lm", se = FALSE, color = "black") +  # Adds a linear regression line
  #facet_wrap(~ Id, scales = "fixed", ncol = 5) #if want hiwahiwalay
  labs(
    title="Total Minutes Asleep vs Total Steps",
    caption="Period: 04/12/2016 to 05/12/2016 Daily"
  ) +
  theme_minimal()
```
####Very Active Minutes vs Sedentary Minutes
```{r}
ggplot(data = daily_merged_B, aes(x = VeryActiveMinutes, y = SedentaryMinutes, color=factor(Id))) +
  geom_point() +
  geom_smooth(method="lm", se = FALSE, color = "black") +
  labs(
    title = "Very Active Minutes vs Sedentary Minutes",
    caption ="Period 04.12.2016 to 05.12.2016 Daily"
  ) +
  theme_minimal()
```
####Activity_Type Minutes boxplot
```{r}
activityMins_data <- daily_merged_B %>%
  select(VeryActiveMinutes,FairlyActiveMinutes,LightlyActiveMinutes,SedentaryMinutes)

# Reshape data from wide to long format
library(tidyr)
activity_data_long <- activityMins_data %>%
  gather(key = "ActivityType", value = "Minutes")

activity_data_long$ActivityType <- factor(activity_data_long$ActivityType,
                                          levels = c("VeryActiveMinutes", "FairlyActiveMinutes", "LightlyActiveMinutes", "SedentaryMinutes"))

# Create the boxplot
ggplot(activity_data_long, aes(x = ActivityType, y = Minutes)) +
  geom_boxplot() +
  stat_summary(fun.data = "mean_sdl", geom = "text", 
               aes(label = paste("Mean: ", round(..y.., 1))),
               vjust = -1.5, size = 3) +  # Add mean label
  stat_summary(fun = "median", geom = "text", 
               aes(label = paste("Median: ", round(..y.., 1))),
               vjust = 1.5, size = 3) +  # Add median label
  theme_minimal() +
  labs(title = "Activity Minutes Boxplot",
       x = "Activity Type",
       y = "Minutes",
       caption="Period: 04.12.2016 to 05.12.2016 Daily")
```
####Activity_Type Distance boxplot
```{r}
activityDist_data <- daily_merged_B %>%
  select(VeryActiveDistance,ModeratelyActiveDistance,LightActiveDistance,SedentaryActiveDistance)

# Reshape data from wide to long format
library(tidyr)
activityD_data_long <- activityDist_data %>%
  gather(key = "ActivityType", value = "KM")

activityD_data_long$ActivityType <- factor(activityD_data_long$ActivityType,
                                          levels = c("VeryActiveDistance","ModeratelyActiveDistance","LightActiveDistance","SedentaryActiveDistance"))

# Create the boxplot
ggplot(activityD_data_long, aes(x = ActivityType, y = KM)) +
  geom_boxplot() +
  stat_summary(fun.data = "mean_sdl", geom = "text", 
               aes(label = paste("Mean: ", round(..y.., 1))),
               vjust = -1.5, size = 3) +  # Add mean label
  stat_summary(fun = "median", geom = "text", 
               aes(label = paste("Median: ", round(..y.., 1))),
               vjust = 1.5, size = 3) +  # Add median label
  theme_minimal() +
  labs(title = "Activity Distance Boxplot",
       x = "Activity Type",
       y = "KM",
       caption="Period: 04.12.2016 to 05.12.2016 Daily")
```

####Total Distance vs Total Steps
```{r}
ggplot(data = daily_merged_B, aes(x = TotalDistance, y = TotalSteps, color=factor(Id))) +
  geom_point() +
  geom_smooth(method="lm", se = FALSE, color = "black") +
  labs(
    title = "Total Distance (km) vs Total Steps",
    caption ="Period 04.12.2016 to 05.12.2016 Daily"
  ) +
  theme_minimal()
```

####Total Time Asleep vs Very Active MINUTES Faily Lightly sedentary (one-by-one)
```{r}
ggplot(data = daily_merged_B, aes(x = TotalMinutesAsleep, y = SedentaryMinutes, color=factor(Id))) + #replace y axis
  geom_point() +
  geom_smooth(method="auto", se = FALSE, color = "black") +
  labs(
    #title = "Total Time Asleep vs Very Active Minutes",
    #title = "Total Time Asleep vs Faily Active Minutes",
    #title = "Total Time Asleep vs Lightly Active Minutes",
    title = "Total Time Asleep vs Sedentary Minutes",
    caption ="Period 04.12.2016 to 05.12.2016 Daily"
  ) +
  theme_minimal()
```

####Total Time Asleep vs Very Active MINUTES, Faily, Lightly, sedentary (in one)
```{r}
activityM_long <- daily_merged_B %>%
  select(Id, TotalMinutesAsleep, VeryActiveMinutes, FairlyActiveMinutes, LightlyActiveMinutes, SedentaryMinutes) %>%
  pivot_longer(cols = c(VeryActiveMinutes, FairlyActiveMinutes, LightlyActiveMinutes, SedentaryMinutes),
               names_to = "ActivityType", values_to = "ActivityMinutes")

# Plot with reshaped data
ggplot(data = activityM_long, aes(x = TotalMinutesAsleep, y = ActivityMinutes, color = ActivityType)) +
  geom_point( alpha = 0.4) +
  geom_smooth(method = "auto", aes(color = ActivityType), size = 1.25, se = FALSE) +  # Color code the smoothed lines
  labs(
    title = "Total Time Asleep vs Activity Minutes by Type",
    caption = "Period 04.12.2016 to 05.12.2016 Daily",
    x = "Total Minutes Asleep",
    y = "Activity Minutes"
  ) +
  theme_minimal()
```

####Total Time Asleep vs Very Active DISTANCE Faily Lightly sedentary (one-by-one)
```{r}
ggplot(data = daily_merged_B, aes(x = TotalMinutesAsleep, y = VeryActiveDistance, color=factor(Id))) + #replace y axis
  geom_point() +
  geom_smooth(method="auto", se = FALSE, color = "black") +
  labs(
    title = "Total Time Asleep vs Very Active Distance (km)",
    #title = "Total Time Asleep vs Moderately Active Distance (km)",
    #title = "Total Time Asleep vs Lightly Active Distance (km)",
    #title = "Total Time Asleep vs Sedentary Active Distance (km)",
    caption ="Period 04.12.2016 to 05.12.2016 Daily"
  ) +
  theme_minimal()
```
####Total Time Asleep vs Very Active DISTANCE, Faily, Lightly, sedentary (in one)
```{r}
activityD_long <- daily_merged_B %>%
  select(Id, TotalMinutesAsleep, VeryActiveDistance, ModeratelyActiveDistance, LightActiveDistance, SedentaryActiveDistance) %>%
  pivot_longer(cols = c(VeryActiveDistance, ModeratelyActiveDistance, LightActiveDistance, SedentaryActiveDistance),
               names_to = "ActivityTypeDistance", values_to = "ActivityDistance")

# Plot with reshaped data
ggplot(data = activityD_long, aes(x = TotalMinutesAsleep, y = ActivityDistance, color = ActivityTypeDistance)) +
  geom_point( alpha = 0.4) +
  geom_smooth(method = "auto", aes(color = ActivityTypeDistance), size = 1.25, se = FALSE) +  # Color code the smoothed lines
  labs(
    title = "Total Time Asleep vs Activity Distance by Type",
    caption = "Period 04.12.2016 to 05.12.2016 Daily",
    x = "Total Minutes Asleep",
    y = "Activity Distance (km)"
  ) +
  theme_minimal()
```


####Relationship betweeen total steps and total calories
```{r}
ggplot(data=daily_merged_B, aes(x=Calories, y=TotalSteps, color=factor(Id)))+
  geom_point()+
  geom_smooth(method = "lm", se = FALSE, color = "black") +  # Adds a linear regression line
  #facet_wrap(~ Id, scales = "fixed", ncol = 5) #if want hiwahiwalay
  labs(
    title="Total Calories Burned vs Total Steps",
    caption="Period: 04/12/2016 to 05/12/2016 Daily",
    x = "Total Calories Burned"
  ) +
  theme_minimal()
```
####Total calories vs Very Active MINUTES Faily Lightly sedentary (one-by-one)
```{r}
ggplot(data = daily_merged_B, aes(x = Calories, y = VeryActiveMinutes, color=factor(Id))) + #replace y axis
  geom_point() +
  geom_smooth(method="auto", se = FALSE, color = "black") +
  labs(
    title = "Total Calories Burned vs Very Active Minutes",
    #title = "Total Calories Burned vs Faily Active Minutes",
    #title = "Total Calories Burned vs Lightly Active Minutes",
    #title = "Total Calories Burned vs Sedentary Minutes",
    caption ="Period 04.12.2016 to 05.12.2016 Daily",
    x = "Total Calories Burned"
  ) +
  theme_minimal()
```

####Total calories vs Very Active MINUTES, Faily, Lightly, sedentary (in one)
```{r}
activityCalM_long <- daily_merged_B %>%
  select(Id, Calories, VeryActiveMinutes, FairlyActiveMinutes, LightlyActiveMinutes, SedentaryMinutes) %>%
  pivot_longer(cols = c(VeryActiveMinutes, FairlyActiveMinutes, LightlyActiveMinutes, SedentaryMinutes),
               names_to = "ActivityType", values_to = "ActivityMinutes")

# Plot with reshaped data
ggplot(data = activityCalM_long, aes(x = Calories, y = ActivityMinutes, color = ActivityType)) +
  geom_point( alpha = 0.1) +
  geom_smooth(method = "auto", aes(color = ActivityType), size = 1.25, se = FALSE) +  # Color code the smoothed lines
  labs(
    title = "Total Calories Burned vs Activity Minutes by Type",
    caption = "Period 04.12.2016 to 05.12.2016 Daily",
    x = "Total Calories Burned",
    y = "Activity Minutes"
  ) +
  theme_minimal()
```

####Total CALORIES vs Very Active DISTANCE Faily Lightly sedentary (one-by-one)
```{r}
ggplot(data = daily_merged_B, aes(x = Calories, y = SedentaryActiveDistance, color=factor(Id))) + #replace y axis
  geom_point() +
  geom_smooth(method="auto", se = FALSE, color = "black") +
  labs(
    #title = "Total Calories Burned vs Very Active Distance (km)",
    #title = "Total Calories Burned vs Moderately Active Distance (km)",
    #title = "Total Calories Burned vs Lightly Active Distance (km)",
    title = "Total Calories Burned vs Sedentary Active Distance (km)",
    caption ="Period 04.12.2016 to 05.12.2016 Daily",
    x = "Total Calories Burned"
  ) +
  theme_minimal()
```
####Total CALORIES vs Very Active DISTANCE, Faily, Lightly, sedentary (in one)
```{r}
activitycalD_long <- daily_merged_B %>%
  select(Id, Calories, VeryActiveDistance, ModeratelyActiveDistance, LightActiveDistance, SedentaryActiveDistance) %>%
  pivot_longer(cols = c(VeryActiveDistance, ModeratelyActiveDistance, LightActiveDistance, SedentaryActiveDistance),
               names_to = "ActivityTypeDistance", values_to = "ActivityDistance")

# Plot with reshaped data
ggplot(data = activitycalD_long, aes(x = Calories, y = ActivityDistance, color = ActivityTypeDistance)) +
  geom_point( alpha = 0.1) +
  geom_smooth(method = "auto", aes(color = ActivityTypeDistance), size = 1.25, se = FALSE) +  # Color code the smoothed lines
  labs(
    title = "Total Calories Burned vs Activity Distance by Type",
    caption = "Period 04.12.2016 to 05.12.2016 Daily",
    x = "Total Calories Burned",
    y = "Activity Distance (km)"
  ) +
  theme_minimal()
```

##HOURLY
###HOURLY-LOAD
```{r}
hourly_calories_B <- read.csv("D:/Users/rjlde/Documents/+STUDY FILES+/2025 CASE STUDIES/CASE STUDY 2/RPROJ/CSV/HOURLY/hourlyCalories_merged_4.12.16-5.12.16.csv", stringsAsFactors = FALSE)
hourly_Intensities_B <- read.csv("D:/Users/rjlde/Documents/+STUDY FILES+/2025 CASE STUDIES/CASE STUDY 2/RPROJ/CSV/HOURLY/hourlyIntensities_merged_4.12.16-5.12.16.csv", stringsAsFactors = FALSE)
hourly_steps_B <- read.csv("D:/Users/rjlde/Documents/+STUDY FILES+/2025 CASE STUDIES/CASE STUDY 2/RPROJ/CSV/HOURLY/hourlySteps_merged_4.12.16-5.12.16.csv", stringsAsFactors = FALSE)
```
Hourly Merge
```{r}
# First merge calories and intensities
hourly_activity <- merge(hourly_calories_B, 
                         hourly_Intensities_B, 
                         by = c("Id", "ActivityHour"),
                         all = TRUE)

# Then merge with steps
hourly_activity <- merge(hourly_activity, 
                         hourly_steps_B, 
                         by = c("Id", "ActivityHour"),
                         all = TRUE)

# Rename the datetime column
hourly_activity <- hourly_activity %>%
  rename(DateHour = ActivityHour)

write.csv(hourly_activity, "D:/Users/rjlde/Documents/+STUDY FILES+/2025 CASE STUDIES/CASE STUDY 2/RPROJ/CSV/CLEANED/hourly_activity_merged.csv", row.names = FALSE)
```
####HOURLY CLEAN
head()
```{r}
head(hourly_activity)
#Findings - ActivityDate is chr not datetime format
```
Number of Unique participants:
```{r}
n_distinct(hourly_activity)
```
####Clean ->
remove duplicated rows:
```{r}
hourly_activity <- hourly_activity[!duplicated(hourly_activity), ]
#no duplicated values found
```
Handle missing values:
```{r}
colSums(is.na(hourly_activity))
#no missing values found
```
Convert data types:
```{r}
library(lubridate)

# Convert DateHour to POSIXct format (for "MM/DD/YYYY h:mm:ss AM/PM")
hourly_activity$DateHour <- mdy_hms(hourly_activity$DateHour)

# Check the result
head(hourly_activity$DateHour)
```

Quick Summary Stats:
```{r}
summary_hourly_activity <- hourly_activity %>%
    summary()

print(summary_hourly_activity)
```
SAVE Quick Summary Stats:
```{r}
write.csv(summary_hourly_activity, "D:/Users/rjlde/Documents/+STUDY FILES+/2025 CASE STUDIES/CASE STUDY 2/RPROJ/CSV/CLEANED/summary_hourly_activity_merged.csv", row.names = FALSE)
```

###HOURLY-EXPLORATION & VISUALIZATION
####Separate Date and Time
```{r}
hourly_activity_timed <- hourly_activity %>%
  mutate(
    Date = as.Date(DateHour),   # Extract the date part
    Time = format(DateHour, "%H:%M:%S")   # Extract the time part (in HH:MM:SS format)
  )

hourly_activity_timed$Time <- as_hms(hourly_activity_timed$Time)
```
check
```{r}
# Check the result
head(hourly_activity_timed)
```
####PER HOUR total STEPS
```{r}
ggplot(hourly_activity_timed, aes(x = Time, y = StepTotal, color = factor(Id))) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "auto", size = 1.25, color = "black", se = FALSE) +
  labs(
    title = "Total Steps by Time of Day",
    x = "Time",
    y = "Total Steps",
    color = "User ID",
    caption = "Period: 4.12.2016 to 5.12.2016 Hourly"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1)  # Rotate x-axis labels by 90 degrees
  )
```

####PER HOUR total INTENSITIES
```{r}
ggplot(hourly_activity_timed, aes(x = Time, y = TotalIntensity, color = factor(Id))) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "auto", size = 1.25, color = "black", se = FALSE) +
  labs(
    title = "Total Activity Intensity by Time of Day",
    x = "Time",
    y = "Total Intensity",
    color = "User ID",
    caption = "Period: 4.12.2016 to 5.12.2016 Hourly"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1)  # Rotate x-axis labels by 90 degrees
  )
```

####PER HOUR total CALORIES
```{r}
ggplot(hourly_activity_timed, aes(x = Time, y = Calories, color = factor(Id))) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "auto", size = 1.25, color = "black", se = FALSE) +
  labs(
    title = "Total Calories Burned by Time of Day",
    x = "Time",
    y = "Total Calories",
    color = "User ID",
    caption = "Period: 4.12.2016 to 5.12.2016 Hourly"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1)  # Rotate x-axis labels by 90 degrees
  )
```


###MINUTE - LOAD HEART RATE
```{r}
secs_heartRate <- read.csv("D:/Users/rjlde/Documents/+STUDY FILES+/2025 CASE STUDIES/CASE STUDY 2/RPROJ/CSV/SECOND/heartrate_seconds_merged_4.12.16-5.12.16.csv", stringsAsFactors = FALSE)
```

####MINUTE CLEAN
head()
```{r}
head(secs_heartRate)
#Findings - Time is chr not datetime format
```

remove duplicated rows:
```{r}
secs_heartRate <- secs_heartRate[!duplicated(secs_heartRate), ]
#no duplicated values found
```
Handle missing values:
```{r}
colSums(is.na(secs_heartRate))
#no missing values found
```
Convert data types:
```{r}
library(lubridate)

# Convert Time to POSIXct format (for "MM/DD/YYYY hh:mm:ss AM/PM")
secs_heartRate$Time <- mdy_hms(secs_heartRate$Time)

# Check the result
head(secs_heartRate)
```
```{r}
n_distinct(secs_heartRate$Id)
```
####add column taking hours
```{r}
secs_heartRate$Hour <- format(secs_heartRate$Time, "%H")
```

```{r}
head(secs_heartRate)
```

group by hour and user id
```{r}
avg_hrtRate_per_user <- secs_heartRate %>%
  group_by(Id, Hour) %>%
  summarise(avg_heart_rate = mean(Value, na.rm = TRUE)) %>%
  ungroup()
```
####PLOT heart rate by time-of-day
```{r}
ggplot(avg_hrtRate_per_user, aes(x = as.numeric(Hour), y = avg_heart_rate, color=factor(Id))) +
  geom_point(alpha = 0.5) +
  geom_smooth(aes(group = 1), method = "loess", se = FALSE, color = "black", size = 1.2) +
  scale_x_continuous(breaks = 0:23) +
  labs(title = "Average Heart Rate by Time-of-Day",
       x = "Hour of Day",
       y = "Average Heart Rate") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```


